---
title: "Examination of Poverty Rates Before and After Housing Cost Across UK Regions: England, Wales, Scotland and Northern Ireland"
author: "Cynthia Banjo"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true       
    number_sections: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Setup and Libraries

```{r}

library(tidyverse)
library(readxl)
library(janitor)  
library(haven)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

## Import and Explore Data

```{r}
hbai1821 <- read_sav("i1821e_2324prices.sav") %>% clean_names()
hbai2124 <- read_sav("i2124e_2324prices.sav") %>% clean_names()
hbai_0912 <- read_sav("i0912e_2324prices.sav") %>% clean_names()

names(hbai_0912)
colnames(hbai1821)
colnames(hbai2124)
```

## Combine Datasets

```{r}
hbai_all <- bind_rows(hbai1821, hbai2124)
glimpse(hbai_all)
summary(hbai_all)
```

## Data Cleaning and Column Selection

```{r}
hbai_clean <- hbai_all %>%
  select(year, country,
         low50ahc, low60ahc, low70ahc, entinchh,
         eqoahchh, eqobhchh,
         low50bhc, low60bhc, low70bhc,
         mdch, mdpn, mdwa) %>%
  mutate(year = as.integer(year))

hbai_0912_selected <- hbai_0912 %>%
  select(
    ahcyrdef, bhcyrdef, year,
    entinchh, low50ahc, low60ahc, low70ahc,
    low50bhc, low60bhc, low70bhc, gvtregn
  )
```

## Establish Baseline and Compute Absolute Poverty

```{r}
baseline_2010 <- hbai_0912_selected %>%
  filter(year == 17) %>%
  summarise(median_2010 = median(entinchh, na.rm = TRUE)) %>%
  pull(median_2010)

absolute_threshold_2010 <- 0.6 * baseline_2010
absolute_threshold_2010
```

```{r}
absolute_poverty <- hbai_0912_selected %>%
  group_by(year) %>%
  summarise(
    median_income = median(entinchh, na.rm = TRUE),
    abs_poverty_rate = mean(entinchh < absolute_threshold_2010, na.rm = TRUE)
  )
absolute_poverty
```

## Poverty Before Housing Costs (BHC)

```{r}
absolute_threshold_bhc <- hbai_0912_selected %>%
  filter(year == 17) %>%
  summarise(median_income_2010 = median(entinchh, na.rm = TRUE)) %>%
  pull(median_income_2010)

poverty_bhc_uk <- hbai_clean %>%
  group_by(year) %>%
  summarise(
    rel_poverty_60 = mean(low60bhc, na.rm = TRUE),
    deep_poverty_50 = mean(low50bhc, na.rm = TRUE),
    abs_poverty_2010 = mean(entinchh < absolute_threshold_bhc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(rel_poverty_60, deep_poverty_50, abs_poverty_2010),
    names_to = "poverty_type", values_to = "rate"
  ) %>%
  mutate(financial_year = case_when(
    year == 25 ~ "2018/19",
    year == 26 ~ "2019/20",
    year == 27 ~ "2020/21",
    year == 28 ~ "2021/22",
    year == 29 ~ "2022/23",
    year == 30 ~ "2023/24"
  ))
```

```{r fig.width=8, fig.height=5}
ggplot(poverty_bhc_uk, aes(x = financial_year, y = rate, color = poverty_type, group = poverty_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(
    values = c("deep_poverty_50" = "#e41a1c",
               "rel_poverty_60" = "#377eb8",
               "abs_poverty_2010" = "#4daf4a"),
    labels = c("Deep Poverty (50% BHC)",
               "Relative Poverty (60% BHC)",
               "Absolute Poverty (2010 Threshold)")
  ) +
  labs(
    title = "UK Poverty Trend Before Housing Costs (2018/19–2023/24)",
    subtitle = "Proportion below 50%, 60%, and 2010/11 threshold",
    x = "Financial Year", y = "Poverty Rate (%)",
    color = "Measure"
  ) +
  theme_minimal(base_size = 14)
```

## Poverty After Housing Costs (AHC)

```{r}
poverty_ahc_1824 <- hbai_clean %>%
  group_by(year) %>%
  summarise(
    median_income_ahc = median(entinchh, na.rm = TRUE),
    relative_poverty_60 = mean(entinchh < 0.6 * median(entinchh, na.rm = TRUE), na.rm = TRUE),
    deep_poverty_50 = mean(entinchh < 0.5 * median(entinchh, na.rm = TRUE), na.rm = TRUE),
    absolute_poverty_2010 = mean(entinchh < absolute_threshold_2010, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(financial_year = case_when(
    year == 25 ~ "2018/19",
    year == 26 ~ "2019/20",
    year == 27 ~ "2020/21",
    year == 28 ~ "2021/22",
    year == 29 ~ "2022/23",
    year == 30 ~ "2023/24"
  )) %>%
  pivot_longer(cols = c(relative_poverty_60, deep_poverty_50, absolute_poverty_2010),
               names_to = "poverty_type", values_to = "rate")
```

```{r fig.width=8, fig.height=5}
ggplot(poverty_ahc_1824, aes(x = financial_year, y = rate, color = poverty_type, group = poverty_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(
    values = c("relative_poverty_60" = "#377eb8",
               "deep_poverty_50" = "#e41a1c",
               "absolute_poverty_2010" = "#4daf4a"),
    labels = c("Relative Poverty (60% AHC)",
               "Deep Poverty (50% AHC)",
               "Absolute Poverty (2010 Threshold)")
  ) +
  labs(
    title = "UK Poverty Trend After Housing Costs (2018/19–2023/24)",
    x = "Financial Year", y = "Poverty Rate (%)",
    color = "Measure"
  ) +
  theme_minimal(base_size = 14)
```

## BHC vs AHC Comparison (UK)

```{r}
poverty_compare <- hbai_clean %>%
  group_by(year) %>%
  summarise(
    low60ahc_mean = mean(low60ahc, na.rm = TRUE),
    low60bhc_mean = mean(low60bhc, na.rm = TRUE)
  ) %>%
  mutate(financial_year = case_when(
    year == 25 ~ "2018/19",
    year == 26 ~ "2019/20",
    year == 27 ~ "2020/21",
    year == 28 ~ "2021/22",
    year == 29 ~ "2022/23",
    year == 30 ~ "2023/24"
  )) %>%
  pivot_longer(cols = c(low60ahc_mean, low60bhc_mean),
               names_to = "measure", values_to = "rate")
```

```{r fig.width=8, fig.height=5}
ggplot(poverty_compare, aes(x = financial_year, y = rate, color = measure, group = measure)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(
    values = c("low60ahc_mean" = "#1f77b4", "low60bhc_mean" = "#2ca02c"),
    labels = c("After Housing Costs (AHC)", "Before Housing Costs (BHC)")
  ) +
  labs(
    title = "Comparison of Poverty Rates (BHC vs AHC)",
    subtitle = "UK Poverty Trends (2018/19–2023/24)",
    x = "Financial Year", y = "Poverty Rate (%)"
  ) +
  theme_minimal(base_size = 14)
```


## Country-Level Comparisons
```{r}
# Map country names to the country codes

hbai_clean <- hbai_clean %>%
  mutate(country_name = case_when(
    country == 1 ~ "England",
    country == 2 ~ "Wales",
    country == 3 ~ "Scotland",
    country == 4 ~ "Northern Ireland",
    TRUE ~ "Unknown"
  ))

### BHC vs ABC 2018/2019
poverty_compare_2018 <- hbai_clean %>%
  filter(year == 25) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60ahc, na.rm = TRUE),
    AHC = mean(low60bhc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")
### Visualization 3.2 BHC vs AHC 2018/2019
ggplot(poverty_compare_2018, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC (2018/19)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

###BHC vs AHC for 2019/2020

poverty_compare_2019 <- hbai_clean %>%
  filter(year == 26) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60bhc, na.rm = TRUE),
    AHC = mean(low60ahc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")

###Visualization 2.3: Poverty BHC vs AHC 2019/20

ggplot(poverty_compare_2019, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC 2019/20)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

###BHC vs AHC for 2020/2021

poverty_compare_2020 <- hbai_clean %>%
  filter(year == 27) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60bhc, na.rm = TRUE),
    AHC = mean(low60ahc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")

###Visualization 2.4: BHC vs AHC  2020/21

ggplot(poverty_compare_2020, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC (2020/21)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

### BHC vs AHC for 2021/2022

poverty_compare_2021 <- hbai_clean %>%
  filter(year == 28) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60bhc, na.rm = TRUE),
    AHC = mean(low60ahc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")

###Visualization 2.5 BHC vs AHC 2021/22

ggplot(poverty_compare_2021, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC (2021/22)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

#BHC vs AHC for 2022/2023

poverty_compare_2022 <- hbai_clean %>%
  filter(year == 29) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60bhc, na.rm = TRUE),
    AHC = mean(low60ahc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")

#Visualization 2.6: HC vs AHC 2022/23

ggplot(poverty_compare_2022, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC (2022/23)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

#BHC vs AHC 2023/2024

poverty_compare_2023 <- hbai_clean %>%
  filter(year == 30) %>%   # 2018/19
  group_by(country_name) %>%
  summarise(
    BHC = mean(low60bhc, na.rm = TRUE),
    AHC = mean(low60ahc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(BHC, AHC),
               names_to = "Type",
               values_to = "Rate")

#Visualization 2.7: BHC vs AHC 2023/2024

ggplot(poverty_compare_2023, aes(x = country_name, y = Rate, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = scales::percent(Rate, accuracy = 1)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  scale_fill_manual(
    values = c(
      "BHC" = "#377eb8",
      "AHC" = "#e41a1c"
    ),
    labels = c(
      "BHC" = "Before Housing Costs",
      "AHC" = "After Housing Costs"
    )
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 0.5)) +
  labs(
    title = "Comparison of Poverty Rates — BHC vs AHC (2023/24)",
    subtitle = "Proportion below 60% of median (BHC/AHC), by country",
    x = "Country",
    y = "Poverty Rate (%)",
    fill = "Measure"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## Conclusion

- Relative poverty remained relatively stable.
- Absolute poverty improved over time.
- AHC poverty is consistently higher than BHC poverty.


